---
title: "Survival data challenge, starting kit"
author: "Florent Chuffart"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---


# Purpose

The challenge is hosted on the codalab polatform:

https://competitions.codalab.org/competitions/21360?secret_key=1b0893b2-c695-4c9b-ae33-334ac7a056ac


```{r, echo=FALSE, eval=TRUE}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>", fig.width=9, fig.height=6, eval=TRUE, echo=TRUE, results="hide")
require(survival)
scurve = function (ss, v, colors = c("deepskyblue", "black", "red"), main = "Survival", 
    legend, nb_sign = 3, legend_place = "topright", PLOT_LEGEND = TRUE, 
    cex.leg = 1, censoring, ...) 
{
    idx = !is.na(ss)
    ss = ss[idx]
    if (missing(v)) {
        v = rep(1, sum(idx))
    }
    else {
        v = v[idx]
    }
    if (!missing(censoring)) {
        date = ss[, 1]
        state = ss[, 2]
        idx = !is.na(date) & date > censoring
        date[idx] = censoring
        state[idx] = 0
        ss = survival::Surv(date, state)
    }
    if (length(unique(v)) == 1) {
        cox_results = 1
    }
    else {
        cox_results = coxres(ss, as.character(v))
    }
    pv = cox_results[1]
    if (pv < 1e-100) {
        pvt = "<1e-100"
    }
    else {
        pvt = format(pv, digits = 3, scientific = TRUE)
    }
    sf = survival::survfit(ss ~ v)
    levels = length(na.omit(unique(v)))
    col = colorRampPalette(colors)(levels)
    main = paste(main, " p_cox=", signif(as.numeric(pvt), nb_sign), 
        sep = "")
    plot(sf, col = col, main = main, ...)
    tab = table(v)
    tab = tab[tab != 0]
    if (PLOT_LEGEND) {
        if (missing(legend)) {
            if ("breaks" %in% names(attributes(v))) {
                b = signif(attr(v, "breaks"), 3)
                legend = paste("[", b[1:(length(b) - 1)], ",", 
                  b[2:length(b)], c(rep("[", length(b) - 2), 
                    "]"), sep = "")
            }
            else {
                legend = names(tab)
            }
        }
        legend = paste(legend, " (", tab, ")", sep = "")
        legend(legend_place, legend = legend, col = col, pch = 3, 
            lty = 1, cex = cex.leg)
    }
    return(cox_results)
}

coxres = function (ss, v) 
{
    f = suppressWarnings(survival::coxph(ss ~ v))
    sf = summary(f)
    pvcox = sf$logtest[3]
    tf = survival::cox.zph(f)
    tf = tf$table
    pvhz = tf[dim(tf)[1], 3]
    hr = sf$conf.int[1, 1]
    hrlb = sf$conf.int[1, 3]
    hrub = sf$conf.int[1, 4]
    res = c(pvcox, pvhz, hrlb, hr, hrub)
    names(res) = c("pvcox", "pvhz", "hrlb", "hr", "hrub")
    return(res)
}
```

# References

TCGA Study Abbreviations:

https://gdc.cancer.gov/resources-tcga-users/tcga-code-tables/tcga-study-abbreviations



# Loading data

```{r}
exp_grp = readRDS(file="challenge_ssd2019_e.rds")
data    = readRDS(file="challenge_ssd2019_d.rds")
pf      = readRDS(file="challenge_ssd2019_p.rds")
```

# Tumoral and non tumoral samples 

```{r}
idx_nt = rownames(exp_grp)[exp_grp$tissue_status == "normal"] 
idx_tu = rownames(exp_grp)[exp_grp$tissue_status == "tumoral"] 
```

# PCA

##  Non tumoral tissues

Guttman effect

```{r}
pca = prcomp(t(data[,idx_nt]), scale=FALSE)
v = pca$sdev * pca$sdev
p = v / sum(v) * 100


col = as.numeric(as.factor(exp_grp[idx_nt,]$tissue_group_level1))
layout(matrix(1:2,1), respect=TRUE)
barplot(p[1:10], ylab="% of var")

i = 1
j = 2
plot(pca$x[,i], pca$x[,j], 
  xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"), 
  ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"),
  col=col
)
legend("topright", pch=16, col=1:length(unique(exp_grp[idx_nt,]$tissue_group_level1)), levels(as.factor(exp_grp[idx_nt,]$tissue_group_level1)))


for (i in c(1,3,5,7)) {
  j = i+1
  plot(pca$x[,i], pca$x[,j], 
    xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"), 
    ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"),
    col=col
  )  
}

```

## Tumoral tissues

```{r}
pca = prcomp(t(data[,idx_tu]), scale=FALSE)
v = pca$sdev * pca$sdev
p = v / sum(v) * 100


col = as.numeric(as.factor(exp_grp[rownames(pca$x),]$main_gse_number))
layout(matrix(1:2,1), respect=TRUE)
barplot(p[1:10], ylab="% of var")

i = 1
j = 2
plot(pca$x[,i], pca$x[,j], 
  xlab=paste0("PC", i, "(", signif(p[i], 3), "%)"), 
  ylab=paste0("PC", j, "(", signif(p[j], 3), "%)"),
  col=col
)
legend("topright", pch=16, col=1:length(unique(exp_grp[rownames(pca$x),]$main_gse_number)), levels(as.factor(exp_grp[rownames(pca$x),]$main_gse_number)))


```


# Survival

We plot survival according to: 

   - cancer type
  - sex
  - cancer stage 
  - level of expression of ATAD2 gene

```{r label="survival"}
layout(matrix(1:2, 1), respect=TRUE)
scurve(exp_grp$os, exp_grp$main_gse_number, main="cancer")
scurve(exp_grp$os, exp_grp$sex, main="sex")
scurve(exp_grp$os, exp_grp$tnm_stage, main="stage")
scurve(exp_grp$os, data["ATAD2",rownames(exp_grp)] > 10.5, main="ATAD2 over expression")
```


# Export results for submission

Here, we define survival groups according to ATAD2 expression.
Then, we export results for challenge 1 and challenge 2.
The zip files have to be submit on the coda lab interface > Participate > Submit / View Results > Submit

Warning:  

  - You have only **one** opportunity to submit to Challenge 2, to avoid over fitting.
  - The Challenge 2 is only reachable after 5PM.
   

## Challenge 1

```{r}
# Define groups
idx = rownames(exp_grp)[exp_grp$sets == "challenge1"]
grps = data["ATAD2",idx] < 10.5
names(grps) = idx

# Export grps 
saveRDS(grps, "grps.rds")
zip("grps_chal1.zip", "grps.rds")
```

## Challenge 2

```{r}
# Define groups
idx = rownames(exp_grp)[exp_grp$sets == "challenge2"]
grps = data["ATAD2",idx] < 10.5
names(grps) = idx

# Export grps 
saveRDS(grps, "grps.rds")
zip("grps_chal2.zip", "grps.rds")
```

# Session Information

```{r, results="verbatim"}
sessionInfo()
```


